---
title: "Reproducible Research: Peer Assessment 1"
output:   html_document
---
#Author: alexvirt1
#Generation time: `r Sys.time()`

## Loading and preprocessing the data
```{r echo=TRUE}
library(dplyr)
library(ggplot2)

data <- read.csv("activity.csv")
datag <- group_by(data, date)

#mean and median data
datas <- summarise(datag, steps=sum(steps))
datasf <- filter(datas, !is.na(steps))
# filter NA data and group by interval column
datagf <- data %>% filter(!is.na(steps)) %>% group_by(interval)
#create table that shows average steps per each interval accross all days
datagfs <- datagf %>% summarise(stepsa = mean(steps))
#filter out NA and summarise steps by date
datagfd <- data %>% filter(!is.na(steps)) %>% group_by(date) %>% summarise(steps = sum(steps))
#add column with rounded steps for data simulation later
datagfs <- datagfs %>% mutate(stepsr = round(stepsa))

```
## What is mean total number of steps taken per day?
```{r echo=TRUE}
hist(datasf$steps, xlab="Steps per day", main="Average total number of steps per day")
abline(v=mean(datasf$steps),col="red")
```

###Average total number of steps per day = `r sprintf("%4f",mean(datasf$steps))`
###Median total number of steps per day = `r sprintf("%4f",median(datasf$steps))`

## What is the average daily activity pattern?
```{r echo=TRUE}
plot(datagfs$interval, datagfs$stepsa, type="l", xlab="Interval", ylab="Steps", main="Average number of steps taken")
abline(v=datagfs[which.max(datagfs$stepsa),1],col="red")
```

###Interval `r datagfs[which.max(datagfs$stepsa),1]` contains the maximum number of steps across all days.

## Imputing missing values
### Strategy: missing data is replaced by average number of steps is taken within a given time interval

```{r}
# now create sub-table that contains ONLY rows with NA steps
datana <- data %>% filter(is.na(steps))
#perform left join with datagfs to create column with simulated steps
datanasimulated <- left_join(datana, datagfs)
#select only required columns to create a table that look identical to the original one but NA is replaced with mean value
datana1 <- datanasimulated %>% select(date, interval, steps = stepsr)
#now bind rows from the original activity table that filtered out NAs and with a new table that replaced NA with simulated data
datasimulated <- rbind(datana1, datagf)
```

### Total number of missing values (NA) = `r length(datana$steps)`
```{r echo=TRUE}
#calculate average number of steps per interval from simulated data
datasimulatedg <- datasimulated %>% group_by(date)
datassim <- summarise(datasimulatedg, steps=sum(steps))
hist(datassim$steps, xlab="Steps per day", main="Average total number of steps per day - including simulated data")
abline(v=mean(datassim$steps),col="red")
```

###Average total number of steps per day including simulated data = `r sprintf("%4f",mean(datassim$steps))`
###Median total number of steps per day including simulated data = `r sprintf("%4f",median(datassim$steps))`
###Impact of imputing missing data:
###Delta average = `r sprintf("%4f",mean(datassim$steps)-mean(datasf$steps))`
###Delta median = `r sprintf("%4f",median(datassim$steps)-median(datasf$steps))`

## Are there differences in activity patterns between weekdays and weekends?
```{r echo=TRUE}
weekdays1 <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday')
datasteps <- data %>% filter(!is.na(steps))
dataw <- datasteps %>% mutate (weekday = factor(weekdays(as.Date(datasteps$date)) %in% weekdays1, levels=c(FALSE, TRUE), labels=c('weekend', 'weekday')))
datawg <- dataw %>% group_by(interval, weekday) %>% summarise(steps = mean(steps))
ggplot(datawg, aes(x=interval,y=steps)) + geom_line() + facet_grid(weekday ~ .)
```

###The plot definitely shows some difference in activity patterns between weekdays and weekends.